---
title: "Census Tracts"
author: "Jack Xu"
execute:
  echo: FALSE
  warning: FALSE
  message: FALSE
---

```{r}
library(tidyverse) 
data_raw <- read_csv("data/nyc_tracts.csv",
                     col_types = cols(
                       tract = col_character(),
                       race = col_character(),
                       med_income = col_character() 
                     ))
```

```{r}
nyc_data <- data_raw %>%
  separate(tract, into = c("tract_id", "county_state"), sep = "; ", extra = "merge") %>%
  separate(county_state, into = c("county", "state"), sep = "; ", extra = "merge") %>%
  mutate(
    med_income = as.numeric(na_if(med_income, "NA")), # Now med_income is character, so na_if works
    county = str_remove(county, " County") # Clean up county names
  ) %>%
  filter(!is.na(med_income)) # Remove rows where median income is NA
# Create the graphic
ggplot(nyc_data, aes(x = county, y = med_income, fill = race)) +
  geom_boxplot(outlier.shape = NA) + # Use box plots to show distribution, hide outliers for cleaner look
  geom_jitter(aes(color = race), width = 0.2, alpha = 0.6, size = 1.5) + # Add jittered points to show individual data points
  facet_wrap(~ race, scales = "free_y") + # Separate plots by race, with free y-scales for better comparison
  labs(
    title = "Median Income Distribution by Race and County in NYC Census Tracts",
    x = "County",
    y = "Median Income ($)",
    fill = "Race",
    color = "Race"
  ) +
  scale_y_continuous(labels = scales::dollar) + # Format y-axis as currency
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels for readability
    plot.title = element_text(hjust = 0.5, face = "bold"), # Center and bold the title
    strip.text = element_text(face = "bold") # Bold facet titles
  )
```

This R code begins by loading the tidyverse package and reading the nyc_tracts.csv file while specifying that all columns should initially be read as character data. The data is then cleaned and transformed by splitting the tract column into separate tract_id, county, and state columns using separate(). The med_income column, originally character, is converted to numeric, with "NA" strings replaced by actual missing values. The county names are cleaned by removing the word “ County” for simplicity. Rows with missing median income are filtered out to ensure valid data for analysis. Finally, the code creates a detailed boxplot visualization of median income distributions across NYC counties, grouped by race. It enhances readability by adding jittered points to show individual data observations, facets the plot by race with free y-axis scales for better comparison, formats the income axis as dollars, and adjusts text and layout for clarity and presentation quality.
